%
% vim: ts=4 sw=4 et
%
\xchapter{intro}{Introduction}

\xsection{whatis}{What Is NASM?}

The Netwide Assembler, NASM, is an 80x86 and x86-64 assembler designed
for portability and modularity. It supports a range of object file
formats, including Linux and ``*BSD a.out'', ``ELF'', ``COFF'',
``Mach-O'', 16-bit and 32-bit ``OBJ'' (OMF) format, ``Win32'' and
``Win64''. It will also output plain binary files, Intel hex and
Motorola S-Record formats. Its syntax is designed to be simple and
easy to understand, similar to the syntax in the Intel Software
Developer Manual with minimal complexity. It supports all currently
known x86 architectural extensions, and has strong support for macros.

NASM also comes with a set of utilities for handling the ``RDOFF''
custom object-file format.

\xsection{license}{License Conditions}

Please see the file \index{license} ``LICENSE'', supplied as part
of any NASM distribution archive, for the license conditions under
which you may use NASM. NASM is now under the so-called 2-clause
BSD license, also known as the simplified BSD license.

Copyright 1996-2017 the NASM Authors - All rights reserved.

Redistribution and use in source and binary forms, with or without
modification, are permitted provided that the following conditions are
met:

\begin{itemize}
	\item{Redistributions of source code must retain the above copyright
	    notice, this list of conditions and the following disclaimer.}

	\item{Redistributions in binary form must reproduce the above copyright
	    notice, this list of conditions and the following disclaimer in the
	    documentation and/or other materials provided with the distribution.}
\end{itemize}

THIS SOFTWARE IS PROVIDED BY THE COPYRIGHT HOLDERS AND
CONTRIBUTORS "AS IS" AND ANY EXPRESS OR IMPLIED WARRANTIES,
INCLUDING, BUT NOT LIMITED TO, THE IMPLIED WARRANTIES OF
MERCHANTABILITY AND FITNESS FOR A PARTICULAR PURPOSE ARE
DISCLAIMED. IN NO EVENT SHALL THE COPYRIGHT OWNER OR
CONTRIBUTORS BE LIABLE FOR ANY DIRECT, INDIRECT, INCIDENTAL,
SPECIAL, EXEMPLARY, OR CONSEQUENTIAL DAMAGES (INCLUDING, BUT
NOT LIMITED TO, PROCUREMENT OF SUBSTITUTE GOODS OR SERVICES;
LOSS OF USE, DATA, OR PROFITS; OR BUSINESS INTERRUPTION)
HOWEVER CAUSED AND ON ANY THEORY OF LIABILITY, WHETHER IN
CONTRACT, STRICT LIABILITY, OR TORT (INCLUDING NEGLIGENCE OR
OTHERWISE) ARISING IN ANY WAY OUT OF THE USE OF THIS SOFTWARE,
EVEN IF ADVISED OF THE POSSIBILITY OF SUCH DAMAGE.
